Retrofitting Effect Handlers onto OCaml

### 2020-11-11 retro-concurrency.tex
Minor update
Updates

The interaction of non-local control flow with systems programming is quite
subtle~\cite{TFP17}. Consider the following function that uses blocking I/O
functions from the OCaml standard library to copy data from the input channel
|ic| to the output channel |oc|:
\begin{lstlisting}
let rec copy ic oc =
  try output_string oc ((input_line ic) ^ "\n");
      copy ic oc
  with _ -> (close_in ic; close_out oc)
\end{lstlisting}

The function |input_line| raises |End_of_file| exception on reaching the end of input.
Although not apparent, the code is written in a defensive style to handle other
exceptional cases such as the channels being closed externally. Both
|input_line| and |output_string| raise |Sys_error| exception if the channel is
closed. This function cleanly closes the channels on exceptional cases;
|close_*| functions do nothing if the channel is already closed.

@@

The initial configuration for an expression $e$ is
$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment and the
stack are empty. The top-level reductions
  \begin{mathpar}
    \begin{array}{cc}
      \textsc{StepC} ~
      \inferrule{(\term, \env, \fl, \ostack) \cstep \config}
                {\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} &
      \textsc{StepO} ~
      \inferrule{(\term, \env, k, \cstack) \ostep \config}
                {\configa{\term}{\env}{\ostacka{k}{\cstack}} \step \config}
    \end{array}
  \end{mathpar}
can be performed by either by taking a C step $\cstep$ or an OCaml step
$\ostep$. Intuitively, this splits the world into executing C code and OCaml
code.

@@

\begin{figure*}
\begin{minipage}{0.35\linewidth}
  \centering
  \includegraphics[scale=0.45]{figures/fiber}
  \subcaption{Fiber layout}
  \label{fig:fiber}
\end{minipage}
\begin{minipage}{0.64\linewidth}
  \begin{minipage}{\linewidth}
    \centering
    \includegraphics[scale=0.4]{figures/multicore_stack}
    \subcaption{Stack layout for \texttt{meander} example from \S\ref{sec:stack}.}
    \label{fig:mcstack}
    \vspace{2mm}
  \end{minipage}
  %
  \begin{minipage}{0.55\linewidth}
    \begin{lstlisting}
effect E : unit
effect F : unit;;
match (* comp_e *)
  match (* comp_f *)
    (*p1*) perform E (*p3*)
  with | v -> v | effect F kf -> ...
with | v -> v
		 | effect E ke ->
		     (*p2*) continue ke ()
    \end{lstlisting}
		\subcaption{Constructing continuation objects}
		\label{code:effimpl}
  \end{minipage}
  %
  \begin{minipage}{0.44\linewidth}
    \centering
    \includegraphics[scale=0.42]{figures/fiber_handler}
		\subcaption{Program state for code in~\ref{code:effimpl}}
    \label{fig:fiber_handler}
  \end{minipage}
\end{minipage}
\mycaption{Layout of Multicore OCaml effect handlers.}
\end{figure*}

@@

We also present micro benchmarks results in Table~\ref{tab:micro_noeffect}.
Since micro benchmarks magnify micro-architectural optimisations, we also
report the number of instructions executed (obtained using |perf|) along with
time. |exnval| perform 100 million iterations of installing exception handlers
and returning with value. |exnraise| is similar, but raises an exception in
each iteration. |extcall| and |callback| perform 100 million external calls and
callbacks to identity functions. The other micro benchmarks are highly
recursive programs and were taken from ~\ref{Folklore}. For micro benchmarks,
we observed that padding tightloops with a few |nop| instructions, which
changes the loop alignment, makes the code upto 15\% faster. Hence, the
difference in running times under 15\% may not be statistically significant.

@@

\begin{table}
\caption{Micro benchmarks with handlers but no perform. Each entry is the
	slowdown factor ($\times$ times) over its idiomatic implementation in stock OCaml.}
\vspace{-3mm}
{
\begin{tabular}{r c c c c c}
	& \textbf{ack} & \textbf{fib} & \textbf{motzkin} & \textbf{sudan} & \textbf{tak} \\ \hline
	\texttt{MC} 	 	& 10.57 & 15.76 & 16.96 & 12.88 & 13.64 \\
	\texttt{monad} 	& 42.79 & 69.77 & 348.69 & 33.29 & 39.24 \\ \hline
\end{tabular}
}
\label{tab:micro_noperform}
\end{table}

@@

\paragraph{Generators} Generators allow data structures to be traversed on
demand. Many languages including JavaScript and Python provide generators as a
built-in primitive. Using effect handlers (|MC|), given any data structure
(|'a t|) and its iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can
derive its generator function (|val next : unit -> 'a option|). The code is
available in the supplementary material. We evaluate the performance of
traversing a complete binary tree of depth 25 using this generator. This
involves $2^{26}$ stack switches in total. For comparison, we also implemented
a hand-written, selective CPSed~\cite{Nielson01},
defunctionalised~\cite{Danvy01} version (|cps|) and a
concurrency monad (|monad|) version of the generator for the tree. Both |cps|
and |monad| versions are specialised to the binary tree, and come with the
usual caveats of not using the stack for function calls. We observed that the
|cps| version was the fastest thanks to specialisation and hand optimisation.
|MC| version was only 2.76$\times$ slower than |cps| while being a generic
solution, whereas the |monad| version was 8.69$\times$ slower than |cps|.

\paragraph{Chameneos} Chameneos~\cite{Chameneos} is a concurrency game aimed at
measuring context switching and synchronization overheads. Our implementation
uses |MVars| for synchronization. We compare effect handler (|MC|), concurrency
monad (|monad|) and Lwt, a widely used concurrency programming library for
OCaml (|lwt|) versions. We observed that |MC| was the fastest, and |monad|
(|lwt|) was 1.67$\times$ (4.29$\times$) slower than |MC|.

\paragraph{Web server} As one of our primary goals of adding effect handlers to
OCaml is to support direct-style concurrency, we implement a full-fledged
HTTP/1.1 web server using effect handlers (|MC|). We use effect handlers to
implement a user-level threading library, which spawns a thread per request. We
use \texttt{httpaf} library~\cite{httpaf} for HTTP handling, and
\texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
\texttt{libev}. Unlike using effect handlers, Lwt version is written in monadic
style, and does not have a notion of thread per request. For comparison, we
also include a Go version (|go|) that uses \texttt{net/http}~\cite{nethttp}
package.

\begin{figure}
	\begin{minipage}{0.49\linewidth}
		\centering
		\includegraphics[width=\linewidth]{http-benchmarks/throughput.pdf}
		\subcaption{Throughput}
		\label{grf:throughput}
	\end{minipage}
	\begin{minipage}{0.49\linewidth}
		\includegraphics[width=\linewidth]{http-benchmarks/latency.pdf}
		\subcaption{Tail latency}
		\label{grf:latency}
	\end{minipage}
	\mycaption{Web server performance.}
	\vspace{-4mm}
\end{figure}

@@

There are several strategies for implementing effect handlers. Eff~\cite{Eff},
Helium~\cite{Biernacki20}, Frank~\cite{Frank}, Links server
backend~\cite{Hillerstrom20} use an interpreter similar to our operational
semantics to implement effect handlers. Effekt~\cite{Effekt}, Links JavaScript
backend~\cite{Hillerstrom20} and Koka~\cite{Leijen17} use type-directed
selective CPS translation. These language are equipped with an effect system,
which allows compiling pure code in direct-style and effectful code in
continuation passing style. Leijen et al.~\cite{Leijen14} implements effect
handlers as a library in C using stack copying. Since C allows pointers into
the stack, care is taken to ensure that when continuations are resumed, the
constituent frames are restored to the same memory addresses as at the time of
capture. Kiselyov et al.~\cite{Kiselyov18} use an implementation of
multi-prompt delimited continuations as an OCaml library~\cite{Kiselyov12} to
embed Eff language in OCaml. Indeed, Forester et al.~\cite{Forster19} showed
that in an untyped setting, effect handlers, monadic reflection and delimited
control can macro-express each other.

@@

Kawahara et al.~\cite{Kawahara20} implement one-shot effect handlers using
coroutines (a non-local control flow mechanism) as a macro-expressible
translation, and present an embedding in Lua and Ruby. Lua provides asymmetric
coroutines~\cite{Lua} where each coroutine uses its own stack similar to how
each handled computation runs in its own fiber in Multicore OCaml. Similar to
Multicore OCaml, Lua does not allow coroutines to include C frames. It is
useful to note that Lua is an interpreted language unlike OCaml and DWARF
unwind tables are not a concern in Lua.

The interaction on non-local control flow and resources have been studied
extensively. Scheme uses |dynamic-wind| to ensure de-allocation (and
re-allocation) of resources when the non-local control leaves and enters back
into a context. |dynamic-wind| is known to be complex and not quite what one
wants for resource clean up~\cite{Kiselyov,Sitaram03}. The main challenge is distinguishing
non-returning exits from ones that would, and clean up to occur only on
non-returning exit. In Multicore OCaml, we assume that continuations are used
linearly, and hence, no clean up occurs when effects are performed but only when
an exception is thrown, perhaps from the use of |discontinue|. This obviates
the need for |dynamic-wind|. Leijen~\cite{Leijen18} explicitly extends effect
handlers with |finally| clause. Instead, Multicore OCaml |discontinue|
primitive builds on extant defensive coding practices against exceptions to
clean up resource.

Languages with effect handlers typically allow continuations to be resumed more
than once to enable use cases such as backtracking search and probabilistic
programming~\cite{Pyro}. Multicore OCaml has an experimental feature for
explicitly cloning continuations, which makes a copy of each of its constituent
fibers. Multi-shot continuations are difficult to reason about in the presence
of linear resources such as file descriptors. Recall that OCaml programs are
written with the expectation that function calls return \emph{exactly-once}
either normally or exceptionally. With multi-shot continuations, they may
return more than once, leading to errors such as double-freeing malloc'ed
memory. Since our goal is to retrofit effect handlers onto OCaml, we relegate
continuation cloning to the OCaml |Obj| module, which contains unsafe features.

>>>>

The interaction of non-local control flow with systems programming is quite
subtle~\cite{TFP17}. Consider the following function that uses blocking I/O
functions from the OCaml standard library to copy data from the input channel
|ic| to the output channel |oc|:
\begin{lstlisting}
let rec copy ic oc =
  try output_string oc ((input_line ic) ^ "\n");
      copy ic oc
  with End_of_file -> close_in ic; close_out oc
	   | e -> close_in ic; close_out oc; raise e
\end{lstlisting}
The function |input_line| raises |End_of_file| exception on reaching the end of
input, which is handled by the exception handler which closes the channels. The
|close_*| functions do nothing if the channel is already closed. The code is
written in a defensive style to handle other exceptional cases such as the
channels being closed externally. Both |input_line| and |output_string| raise
|Sys_error| exception if the channel is closed. In this case, the channels are
closed and the exception is reraised to communicate the exceptional behaviour
to the caller.

@@

The initial configuration for an expression $e$ is
$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment and the
stack are empty. The top-level reductions
\begin{smathpar}
	\begin{array}{cc}
		\textsc{StepC} ~
		\inferrule{(\term, \env, \fl, \ostack) \cstep \config}
							{\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} &
		\textsc{StepO} ~
		\inferrule{(\term, \env, k, \cstack) \ostep \config}
							{\configa{\term}{\env}{\ostacka{k}{\cstack}} \step \config}
	\end{array}
\end{smathpar}
can be performed by either by taking a C step $\cstep$ or an OCaml step
$\ostep$. Intuitively, this splits the world into executing C code and OCaml
code.

@@

\begin{figure*}
\begin{minipage}{0.35\linewidth}
  \centering
  \includegraphics[scale=0.45]{figures/fiber}
  \subcaption{Fiber layout}
  \label{fig:fiber}
\end{minipage}
\begin{minipage}{0.64\linewidth}
  \begin{minipage}{\linewidth}
    \centering
    \includegraphics[scale=0.4]{figures/multicore_stack}
    \subcaption{Stack layout for \texttt{meander} example from \S\ref{sec:stack}.}
    \label{fig:mcstack}
    \vspace{2mm}
  \end{minipage}
  %
  \begin{minipage}{0.55\linewidth}
    \begin{lstlisting}
effect E : unit
effect F : unit;;
match (* comp_e *)
  match (* comp_f *)
    (*p1*) perform E (*p3*)
  with | v -> v | effect F kf -> ...
with | v -> v
		 | effect E ke ->
		     (*p2*) continue ke ()
    \end{lstlisting}
		\subcaption{Constructing continuation objects}
		\label{code:effimpl}
  \end{minipage}
  %
  \begin{minipage}{0.44\linewidth}
    \centering
    \includegraphics[scale=0.42]{figures/fiber_handler}
		\subcaption{Program state for code in~\ref{code:effimpl}}
    \label{fig:fiber_handler}
  \end{minipage}
\end{minipage}
\vspace{-2mm}
\caption{Layout of Multicore OCaml effect handlers.}
\vspace{-3mm}
\end{figure*}

@@

We also present micro benchmarks results in Table~\ref{tab:micro_noeffect}.
Since micro benchmarks magnify micro-architectural optimisations, we also
report the number of instructions executed (obtained using |perf|) along with
time. |exnval| perform 100 million iterations of installing exception handlers
and returning with value. |exnraise| is similar, but raises an exception in
each iteration. |extcall| and |callback| perform 100 million external calls and
callbacks to identity functions. The other micro benchmarks are highly
recursive programs and were taken from ~\cite{Farvardin20}. For micro benchmarks,
we observed that padding tightloops with a few |nop| instructions, which
changes the loop alignment, makes the code upto 15\% faster. Hence, the
difference in running times under 15\% may not be statistically significant.

@@

\begin{table}
\caption{Micro benchmarks with handlers but no perform. Each entry is the
	slowdown factor ($\times$ times) over its idiomatic implementation in stock OCaml.}
\vspace{-3mm}
{
\begin{tabular}{r c c c c c}
	& \textbf{ack} & \textbf{fib} & \textbf{motzkin} & \textbf{sudan} & \textbf{tak} \\ \hline
	\texttt{MC} 	 	& 10.57 & 15.76 & 16.96 & 12.88 & 13.64 \\
	\texttt{monad} 	& 42.79 & 69.77 & 348.69 & 33.29 & 39.24 \\ \hline
\end{tabular}
}
\label{tab:micro_noperform}
	\vspace{-3mm}
\end{table}

@@

\textbf{\textit{Generators.}} Generators allow data structures to be traversed on
demand. Many languages including JavaScript and Python provide generators as a
built-in primitive. Using effect handlers (|MC|), given any data structure
(|'a t|) and its iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can
derive its generator function (|val next : unit -> 'a option|). The code is
available in the supplementary material. We evaluate the performance of
traversing a complete binary tree of depth 25 using this generator. This
involves $2^{26}$ stack switches in total. For comparison, we also implemented
a hand-written, selective CPSed~\cite{Nielson01},
defunctionalised~\cite{Danvy01} version (|cps|) and a
concurrency monad (|monad|) version of the generator for the tree. Both |cps|
and |monad| versions are specialised to the binary tree, and come with the
usual caveats of not using the stack for function calls. We observed that the
|cps| version was the fastest thanks to specialisation and hand optimisation.
|MC| version was only 2.76$\times$ slower than |cps| while being a generic
solution, whereas the |monad| version was 8.69$\times$ slower than |cps|.

\textbf{\textit{Chameneos.}} Chameneos~\cite{Chameneos} is a concurrency game aimed at
measuring context switching and synchronization overheads. Our implementation
uses |MVars| for synchronization. We compare effect handler (|MC|), concurrency
monad (|monad|) and Lwt, a widely used concurrency programming library for
OCaml (|lwt|) versions. We observed that |MC| was the fastest, and |monad|
(|lwt|) was 1.67$\times$ (4.29$\times$) slower than |MC|.

\textbf{\textit{Web server.}} As one of our primary goals of adding effect handlers to
OCaml is to support direct-style concurrency, we implement a full-fledged
HTTP/1.1 web server using effect handlers (|MC|). We use effect handlers to
implement a user-level threading library, which spawns a thread per request. We
use \texttt{httpaf} library~\cite{httpaf} for HTTP handling, and
\texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
\texttt{libev}. Unlike using effect handlers, Lwt version is written in monadic
style, and does not have a notion of thread per request. For comparison, we
also include a Go version (|go|) that uses \texttt{net/http}~\cite{nethttp}
package.

\begin{figure}
	\begin{minipage}{0.49\linewidth}
		\centering
		\includegraphics[width=\linewidth]{http-benchmarks/throughput.pdf}
		\subcaption{Throughput}
		\label{grf:throughput}
	\end{minipage}
	\begin{minipage}{0.49\linewidth}
		\includegraphics[width=\linewidth]{http-benchmarks/latency.pdf}
		\subcaption{Tail latency}
		\label{grf:latency}
	\end{minipage}
	\vspace{-3mm}
	\caption{Web server performance.}
	\vspace{-4mm}
\end{figure}

@@

There are several strategies for implementing effect handlers. Eff~\cite{Eff},
Helium~\cite{Biernacki20}, Frank~\cite{Frank}, Links server
backend~\cite{Hillerstrom20} use an interpreter similar to our operational
semantics to implement effect handlers. Effekt~\cite{Effekt}, Links JavaScript
backend~\cite{Hillerstrom20} and Koka~\cite{Leijen17} use type-directed
selective CPS translation. These language are equipped with an effect system,
which allows compiling pure code in direct-style and effectful code in
continuation passing style. Leijen et al.~\cite{Leijen14} implement effect
handlers as a library in C using stack copying. Since C allows pointers into
the stack, care is taken to ensure that when continuations are resumed, the
constituent frames are restored to the same memory addresses as at the time of
capture. Kiselyov et al.~\cite{Kiselyov18} use an implementation of
multi-prompt delimited continuations as an OCaml library~\cite{Kiselyov12} to
embed Eff language in OCaml. Indeed, Forester et al.~\cite{Forster19} showed
that in an untyped setting, effect handlers, monadic reflection and delimited
control can macro-express each other.

@@

Kawahara et al.~\cite{Kawahara20} implement one-shot effect handlers using
coroutines (a non-local control flow mechanism) as a macro-expressible
translation, and present an embedding in Lua and Ruby. Lua provides asymmetric
coroutines~\cite{Lua} where each coroutine uses its own stack similar to how
each handled computation runs in its own fiber in Multicore OCaml. Similar to
Multicore OCaml, Lua does not allow coroutines to include C frames. It is
useful to note that Lua is an interpreted language, unlike OCaml, and DWARF
unwind tables are not a concern in Lua.

The interaction on non-local control flow and resources have been studied
extensively. Scheme uses |dynamic-wind| to ensure de-allocation (and
re-allocation) of resources when the non-local control leaves and enters back
into a context. |dynamic-wind| is considered to be complex and not quite what
one wants for resource clean up~\cite{Kiselyov,Sitaram03}. The main challenge
is distinguishing returning exits from a computation closing over resources
from exits that would not return. Resource clean up needs to occur only during
non-returning exit. In Multicore OCaml, we assume that continuations are used
linearly, and hence, no clean up occurs when effects are performed but only
when an exception is thrown, perhaps from the use of |discontinue|. This
obviates the need for |dynamic-wind|.

OCaml does not have a |try/finally| construct available in many programming
languages. In OCaml, alternative standard libraries such as Base~\cite{} and
Core~\cite{} provide mechanisms analogous to Common Lisp |unwind-protect|,
which are in turn implemented using OCaml exceptions. Leijen~\cite{Leijen18}
explicitly extends effect handlers with |finally| clause in Koka. Instead,
Multicore OCaml |discontinue| primitive builds on extant defensive coding
practices against exceptions to clean up resources. This design choice is
crucial to ensure backwards compatibility of legacy systems code with non-local
control flow introduced by effect handlers. Sivaramakrishnan et
al.~\cite{TFP17} describe the interaction of effect handlers and asynchronous
exceptions. This is orthogonal to the contributions of this paper which
focusses on compiler and runtime system support for implementing effect
handlers. Ahman et al.~\cite{Ahman20} present a effect handler core calculus
with asynchrony which might help lay the formal foundations for reconciling
asynchrony in systems and effect handlers.

Languages with effect handlers typically allow continuations to be resumed more
than once to enable use cases such as backtracking search and probabilistic
programming~\cite{Pyro}. Multicore OCaml has an experimental feature for
explicitly cloning continuations, which makes a copy of each of its constituent
fibers. Multi-shot continuations are difficult to reason about in the presence
of linear resources such as file descriptors. Recall that OCaml programs are
written with the expectation that function calls return \emph{exactly-once}
either normally or exceptionally. With multi-shot continuations, they may
return \emph{more than once}, leading to errors such as double-freeing
malloc'ed memory. Since our goal is to retrofit effect handlers onto OCaml, we
relegate continuation cloning to the OCaml |Obj| module, which contains unsafe
features.
